---
title: "Ethan's Project for Fall 2020"
author: "Ethan Tucker"
date: "June 26th, 2020"
output: html_document
---

ATTRIBUTION:
The data I use in this analysis were collected by Andrew Eckstrom of (?). Without this data none of the work that follows would have been possible. I also want to thank Professor Nordmoe for helping me with this project, from connecting me with the data set to all the helpful waychecks!

Potential error: Dataset does not distinguish between sections, so class tracker puts all sections together regardless of whether teaching quality and grade performance varied between sections.

Questions:

1) Does the grade assigned by first Professor predict / influence future performance in major / out of major?

```{r}
Gen_Data %>% 
  filter(`Student Random ID` < 100) %>%
      ggplot(aes(x = Student.Term, y = `GPA Assigned`, color = factor(`Student Random ID`))) +
      geom_point(position = position_jitter(width = 0.2, height = 0.1)) + 
      geom_smooth(method = "lm", se = FALSE)
```


2) Does early exposure to mathematics predict / influence future performance in/out of major? Vice Versa?

3) Do grades generally improve, worsen, or remain constant?

4) Do some teachers have significantly higher Withdraw rates than other professors? (p = .01)
  - Which courses have significantly higher withdraw rates?

5) What is the mean number of different courses a prof teaches? What is the mean number of majors a prof teaches?


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r Initialize}
library(readxl)
library(tidyverse)
library(merTools)
library(lme4)
Easy_Data <- read_excel("/home/local/KNET/k16et01/Homeworks/Research/Stats Research Math, Chemistry, and Biology Grades, Selected Courses.xlsx")
Gen_Data <- read_excel("/home/local/KNET/k16et01/Homeworks/Research/Stats Research XLS.xlsx")
set.seed(1000)
```

ISSUE WITH DATA SET: THE EASY DATA TRACKER DOES NOT DIFFERENTIATE BETWEEN TYPES OF WITHDRAWS. Email Dr. Eckstrom?
ISSUE WITH DATA SET: HARD TO DIFFERENTIATE BETWEEN DIFFERENT SECTIONS Uni



GENDATA ANALYSIS (CHEMISTRY AND MATH STUDENTS)

```{r Gen_Data Semester Numbering, cache = TRUE}
##Fix Gen_Data semester listing. We want them listed numerically from 1 to N.

Number.Semesters <- function(x) {
  x <- x %>% arrange(Semester)
  
  individual.cases <- x %>% pull(Semester)
  
  individual.cases[1]
  
    for(y in seq_along(along.with = individual.cases)){
      if(individual.cases[y] == "2010/FA"){
        individual.cases[y] <- 1}
      else if(individual.cases[y] == "2011/WI"){
        individual.cases[y] <- 2}
      else if(individual.cases[y] == "2011/SU"){
        individual.cases[y] <- 3}
      else if(individual.cases[y] == "2011/FA"){
        individual.cases[y] <- 4}
      else if(individual.cases[y] == "2012/WI"){
        individual.cases[y] <- 5}
      else if(individual.cases[y] == "2012/SU"){
        individual.cases[y] <- 6}
      else if(individual.cases[y] == "2012/FA"){
        individual.cases[y] <- 7}
      else if(individual.cases[y] == "2013/WI"){
        individual.cases[y] <- 8}
      else if(individual.cases[y] == "2013/SU"){
        individual.cases[y] <- 9}
      else if(individual.cases[y] == "2013/FA"){
        individual.cases[y] <- 10}
      else if(individual.cases[y] == "2014/WI"){
        individual.cases[y] <- 11}
      else if(individual.cases[y] == "2014/SU"){
        individual.cases[y] <- 12}
      else if(individual.cases[y] == "2014/FA"){
        individual.cases[y] <- 13}
      else if(individual.cases[y] == "2015/WI"){
        individual.cases[y] <- 14}
      else if(individual.cases[y] == "2015/SU"){
        individual.cases[y] <- 15}
      else if(individual.cases[y] == "2015/FA"){
        individual.cases[y] <- 16}
      else if(individual.cases[y] == "2016/WI"){
        individual.cases[y] <- 17}
      else if(individual.cases[y] == "2016/SU"){
        individual.cases[y] <- 18}
      else if(individual.cases[y] == "2016/FA"){
        individual.cases[y] <- 19}
      else if(individual.cases[y] == "2017/WI"){
        individual.cases[y] <- 20}
      else{stop("There was an invalid Semester name passed to Number.Semesters")}
      }
    x <- mutate(x, Total.Term = individual.cases)
    x
  }
  
Gen_Data <- Number.Semesters(Gen_Data)
```


```{r}
##Total number of professors
Profs.List <- unique(Gen_Data$`Faculty Random ID`)
(n.p <- length(Profs.List))
##Total number of students
Students.List <- unique(Gen_Data$`Student Random ID`)
(n.s <- length(Students.List))
##Total quantity of course entries
n <- length(Gen_Data$`Course Code`)
```

```{r Gen_Data.Grade.Converter} 
##This chunk creates the GPA colunmn in Gen_Data
###Following grade converter obtained and adapted from stack overflow user A5C1D2H2I1M1N2O1R2T1 on the page https://stackoverflow.com/questions/22746508/r-simplifying-code-to-convert-letter-grades-to-numeric-grades on 8/19/2020

convert_grades <- function(x) {
    if (x == "WS") {
        x <- 0
    } else if (x == "WF"){
      x <- 0
    } else if (x == "W"){
      x <- NA
    } else if (x == "WP"){
      x <- NA
    } else if (x == "A") {
        x <- 4
    } else if (x == "A-") {
        x <- 3.7
    } else if (x == "B+") {
        x <- 3.3
    } else if (x == "B") {
        x <- 3
    } else if (x == "B-") {
        x <- 2.7
    } else if (x == "C+") {
        x <- 2.3
    } else if (x == "C") {
        x <- 2
    } else if (x == "C-") {
        x <- 1.7
    } else if (x == "D+") {
        x <- 1.3
    } else if (x == "D") {
        x <- 1
    } else if (x == "D-") {
        x <- 0.7
    } else if (x == "F") {
        x <- 0
    } else {
        x <- NA
    }
  x <- as.double(x)
    return(x)
}


## Fix broken data points
for(i in 1:n){
  if(Gen_Data$Grade[i] == "@D"){
    Gen_Data$Grade[i] <- "D"
  }
  if(Gen_Data$Grade[i] == "@F"){
    Gen_Data$Grade[i] <- "F"
  }
}

##Apply Grade Converter
Ass.GPA <- Gen_Data$Grade
Ass.GPA [] <- sapply(Ass.GPA, convert_grades)
Gen_Data <- mutate(Gen_Data, "GPA Assigned" = as.numeric(Ass.GPA))
```

```{r Gen_Data.Reset.Button}
##Reset Button if below chunk doesn't work
Gen_Data <- dplyr::select(Gen_Data,"Student Random ID":"GPA Assigned")
```

```{r Class Tracker Mean GPA Assigned and Delta, cache=TRUE}
##This chunk creates the Class Tracker, Mean GPA, and Delta columns in Gen_Data. This can be recreated nicely with dplyr, I'll do that if I have time.

##IMPORTANT NOTE: the new Gen_Data variable Class Mean GPA is calculated with na.rm = TRUE, and therefore it is the mean gpa assigned of those students who did not withdraw or take incompletes that did not get resolved.
##Initial Conditions
f <- 1
c <- 1
cntr <- 1
prv.cntr <- 0
Gen_Data <- mutate(Gen_Data, "Class Tracker" = NA, "Class Mean GPA" = NA, "Delta" = NA)
Gen_Data <- arrange(arrange(arrange(Gen_Data,`Faculty Random ID`),Semester),`Course Code`)

while(is.na(Gen_Data$`Class Tracker`[n])){
if(cntr == 4332){
  break
}
  
Current.Course.Name <- Gen_Data$`Course Code`[f]
Current.Semester <- Gen_Data$Semester[f]
Current.Professor <- Gen_Data$`Faculty Random ID`[f]

while(cntr != prv.cntr ){

  if(Gen_Data$`Course Code`[c] == Current.Course.Name && Gen_Data$Semester[c] == Current.Semester && Gen_Data$`Faculty Random ID`[c] == Current.Professor){
    Gen_Data$`Class Tracker`[c] <- cntr
    c <- c+1
    }
  
  else{
    ##Calculate class mean GPA and individual student Delta's
    Class.GPAS <- rep(0,c-f)
    for(z in 2:(c-f)){
      Class.GPAS[z] <- as.double(Gen_Data$"GPA Assigned"[z-1+f])
    }
     class.mean.gpa <- mean(Class.GPAS, na.rm = TRUE)
    
    for(k in f:(c-1)){
      Gen_Data$`Class Mean GPA`[k] <- class.mean.gpa
      Gen_Data$`Delta`[k] <- Gen_Data$"GPA Assigned"[k] - class.mean.gpa
    }
    prv.cntr <- prv.cntr + 1
  }
}

cntr <- cntr + 1

## c-f store to vector one known number of classes
f <- c
}

##There's some jank where this doesn't do the class gpa and delta for the final Class Tracker, so below I tell R to do it manually. This is terrible coding hide your dignity. My one thought was to add a dummy row at the end of Gen_Data and delete it after the loop, because I think R is stopping the loop as soon as the final Class Tracker is assigned, and completes neither Delta nor Class Mean GPA for the final class.
```

```{r}
##Rewrite above code chunk with dplyr if time allows
```

JANK:
start id 139176
end id 139233

```{r Fix final class}
##Fix final class from jank method
f <- 139176
l <- 139233
k <- l - (f-1)
Gen_Data <- arrange(arrange(arrange(Gen_Data,`Faculty Random ID`),Semester),`Course Code`)
Class.GPAS2 <- rep(0,k)
for(i in f:l){
  Class.GPAS2[i-(f-1)] <- as.double(Gen_Data$"GPA Assigned"[i])
  Gen_Data$`Class Tracker`[i] <- 4332 ##CHANGE
}
class.mean.gpa <- mean(Class.GPAS2,na.rm = TRUE)
for(i in f:l){
  Gen_Data$`Class Mean GPA`[i] <- class.mean.gpa
  Gen_Data$Delta[i] <- Gen_Data$"GPA Assigned"[i] - class.mean.gpa
}

```

```{r Initialize Student.Term}
##Here we convert Total.Term into numeric, and create the Student.Term variable which describes a student's term progression through OCC
Gen_Data$Total.Term <- parse_number(Gen_Data$Total.Term)
Gen_Data <- Gen_Data %>%
              arrange(`Student Random ID`, -desc(Total.Term)) %>%
                mutate(Student.Term = 0)

Term.Cntr <- 0
Starting.Student <- Gen_Data$`Student Random ID`[1]
Student.Cntr <- Starting.Student
for(i in 1:n){
  if(Gen_Data$`Student Random ID`[i] == Student.Cntr){
    Term.Cntr <- Term.Cntr + 1
  }
  else{
    Term.Cntr <- 1
    Student.Cntr <- Gen_Data$`Student Random ID`[i]
  }
  Gen_Data$Student.Term[i] <- Term.Cntr
}
```


```{r Department} 
##Create Department variable which describes the department wherein a class was taken.
Gen_Data <- Gen_Data %>% mutate(Department = substr(Gen_Data$`Course Code`,1,3))
```

```{r}
##Create a new data frame that tracks individual students through their years at OCC. First we need to figure out how many classes a student takes in math/chem departments by term.

N.Math.By.Term <- Gen_Data %>% arrange(Term.Number, `Student Random ID`, `Course Code`) %>%
  group_by(`Student Random ID`, Semester, Department) %>%
    count()

N.Math.By.Term %>% filter(n > 1) %>%
  arrange(`Student Random ID`, Semester)

##This dataset can be used to show that students are much more likely to take two math courses simultaneously than two chemistry courses. Additionally, it tells us we need to make a robust points tracking algorithm.
```

```{r}
##Create class size variable for statistical significance tests. There are 5 courses with iffy class sizes: 861, 2636, 3894, 4234, and 3148. The first four of these have n = 1 - they are independent studies. 3148 had n = 3, I'm not sure how that happened. Doing this filtration gives n >= 7, which will allow for safe usage of the t test. Fortunately we did not need to remove almost any cases, so validity of results is basically unaffected.
Sig.Courses <- Gen_Data %>%
                group_by(`Course Code`,`Class Tracker`) %>%
                  count() %>%
                      mutate(class_size = n, Significant = if_else(n > 5, TRUE, FALSE)) %>%
                        dplyr::select(-n) %>%
                         filter(Significant == TRUE)
Gen_Data <- left_join(Sig.Courses, Gen_Data, by = c("Class Tracker", "Course Code")) %>%
              dplyr::select(-Significant)
```

```{r Initialize.Points.Tracker}
##Creates Points Tracker
  
Gen_Data %>%
  group_by(`Class Tracker`) %>% 
    summarise(Prof = head(`Faculty Random ID`, n = 1), Department = substr(head(Gen_Data$`Course Code`, n = 1),1,3))
  
Points.Tracker <- data.frame(Prof.Type = NA, Professors = Profs.List, First.Math.Points = 0, First.Chem.Points = 0, First.Bio.Points = 0, First.Withdraw.Prop = 0,  Total.Math.Points = 0, Total.Chem.Points = 0, Total.Bio.Points = 0,  Total.Withdraw.Prop = 0, Future.Withdraw.Prop = 0)
Points.Tracker <- arrange(Points.Tracker, `Professors`)
View(Points.Tracker)
```


```{r}
##Assigns point values to professors



```




```{r No muli-department faculty}
## This chunk shows that there are no faculty that teach in multiple departments.
counter <- Gen_Data %>%
            group_by(`Faculty Random ID`, Department) %>%
             count() 
length(counter$`Faculty Random ID`) - length(unique(Gen_Data$`Faculty Random ID`))
```

```{r Professor Withdraw Tracker}
##Counts the proportion of withdraws per course that a professor teaches by class size. Compares to expected values.
Withdraws.Data <- Gen_Data  %>%
                    arrange(`Class Tracker`) %>%
                      group_by(`Course Code`, `Faculty Random ID`) %>%
                        filter(is.na(`GPA Assigned`)) %>%
                          count() %>%
                            arrange(`Faculty Random ID`) %>%
                              rename(n_withdraws = n)

          
Total <- Gen_Data  %>%
           arrange(`Class Tracker`) %>%
              group_by(`Course Code`, `Faculty Random ID`) %>%
                count() %>%
                  arrange(`Faculty Random ID`) %>%
                    rename(total_students = n)

Gen_Data_withdraw_prop <- length(Gen_Data$`GPA Assigned`[is.na(Gen_Data$`GPA Assigned`)])/n

Withdraws.Data <- left_join(Withdraws.Data, Total, by = c("Course Code", "Faculty Random ID")) %>%
                  mutate(observed_withdraw_prop = n_withdraws/total_students, expected_withdraws = Gen_Data_withdraw_prop*total_students)


withdraws.model <- lmer(n_withdraws ~ total_students + (1|`Course Code`), data = Withdraws.Data)
summary(withdraws.model)



##Visualization of effect of `Course Code` on n_withdraws
plotREsim(REsim(withdraws.model))

##plot(withdraws.model)
##

withdraw.conf <- confint(withdraws.model, oldNames = FALSE)


Plot.Data1 <- Withdraws.Data %>%
                group_by(`Faculty Random ID`) %>%
                  summarise(total_withdraws = sum(n_withdraws), total_students = sum(total_students))

Plot.Data1 %>%
  ggplot(aes(x = total_withdraws, y = total_students)) +
  geom_point(aes(color = factor(`Faculty Random ID`))) + 
  geom_smooth(color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  theme(legend.position = "none") +
  labs(x = "Total withdraws", y = "Total students", title = "Total withdraws vs. total students, ")  
  

##geom_abline(slope = 1/withdraw.conf[4,1:2], intercept = withdraw.conf[3,1:2], color = "red") +
```



```{r}
##First Course Tracker
Gen_Data <- arrange(Gen_Data)
Gen_Data <- arrange(arrange(Gen_Data,Semester),`Student Random ID`)
f <- 1
l <- 1
cntr <- 1
for(i in 1:n.s){
  ## Picks out each student's classes from Gen Data
  while(Gen_Data$`Student Random ID`[cntr] == Gen_Data$`Student Random ID`[cntr+1]){
    cntr <- cntr + 1
  }
  
  l <- cntr
  
  ##Expresses those classes as a data frame
S.Data <- Gen_Data[f:l,]
  
  
  ##Creates information about the professor to which we assign points
  z <- 1 + l - f
  
  First.Prof <- function(Department){
    m <- length(S.Data$`Student Random ID`)
  }
  
  First.Prof <- S.Data$`Faculty Random ID`[z]
  Prof.Major <- NA
  
  if(startsWith(S.Data$`Course Code`[z],"MAT")){
    Prof.Major <- "MAT"
  }
  else if(startsWith(S.Data$`Course Code`[z],"CHE")){
    Prof.Major <- "CHE"
  }
  
  for(m in 1:n.p){
    if(First.Prof == Profs.List[m]){
      Prof.Position <- m 
    }
  }
  
  
  Sort.Points.First <- function(Major.Points){
    
  }
    
  ## FIRST TEACHER MATH:   Assign math points
    
  ## FIRST TEACHER MATH:   Assign chemistry points
  
  ## FIRST TEACHER MATH:   Assign bio points
  
  ## FIRST TEACHER CHEM:   Assign chem points
  
  ## FIRST TEACHER CHEM:   Assign math points
  
  ## FIRST TEACHER CHEM:   Assign bio points
  
  ## FIRST TEACHER BIO:    Assign bio points
  
  ## FIRST TEACHER BIO:    Assign math points
  
  ## FIRST TEACHER BIO:    Assign chem points
  
  
  
  
  ## Here I check the rest of the profs, and assign them points for OUT OF MAJOR classes.
  
  ## Here I check the rest of the profs, and assign them points for IN MAJOR classes.
  
  
  
  f <- l + 1
}


```

```{r}
##Total Withdraws Tracker

```
















WITH BIOLOGY STUDENTS



```{r}
##Integrate Biology Classes from Easy_Data to Gen_Data
n <- length(Easy_Data$Unique_Student_ID)
n.BIO <- 0
for(i in 1:n){
  if(startsWith(Easy_Data$SUBJ_CODE[i], "BIO")){
    n.BIO <- n.BIO + 1
  }
}
Rows.To.Remove <- rep(4,n)
for(i in 1:n){
  if(!startsWith(Easy_Data$SUBJ_CODE[i],"BIO")){
    Rows.To.Remove[i] <- i
  }
}

Pre.Bio.Data <- Easy_Data[-Rows.To.Remove,]

Pre.Bio.Data <- mutate(rename(Pre.Bio.Data, GPA = GRDE_CODE), Grade = NA)

##Inverse Grade Converter to normalize grade assignments with GenData

##Save non NA grade points as doubles
for(i in 1:length(Pre.Bio.Data$GPA)){
  
if(is.na(Pre.Bio.Data$GPA[i] == FALSE)){
  Pre.Bio.Data$GPA[i] <- as.double(Pre.Bio.Data$GPA[i])
}

}
   
for(i in 1:length(Pre.Bio.Data$Unique_Student_ID)){
  
  G <- Pre.Bio.Data$GPA[i]
  
  if(G == 0.5 || G == 1.5 || G == 2.5 || G == 3.5){
    G <- as.double(G)
    flip <- sample(0:1,1)
    if(flip == 0){
      G <- G - 0.1
    }
    else{
      G <- G + 0.1
    }
  }
  
  if(G >= 0 && G <= 4.1){
    G <- as.double(G)
  }
  
  if(is.character(G)){
  if(G == "I" || G == "W"){
    Pre.Bio.Data$Grade[i] <- "W" 
    Pre.Bio.Data$GPA[i] <- NA
  }
  if(G == "R" || G == "Z"){
   Pre.Bio.Data$Grade[i] <- "F"
   Pre.Bio.Data$GPA[i] <- 0
  }
    G <- NA
  }
  
  
  
  if(is.na(G)){
    
  }
  else if(G >= 3.9){
    Pre.Bio.Data$Grade[i] <- "A"
    Pre.Bio.Data$GPA[i] <- 4.0
  }
  else if(G >= 3.6){
    Pre.Bio.Data$Grade[i] <- "A-"
    Pre.Bio.Data$GPA[i] <- 3.7
  }
  else if(G >= 3.2){
    Pre.Bio.Data$Grade[i] <- "B+"
    Pre.Bio.Data$GPA[i] <- 3.3
  }
  else if(G >= 2.9){
    Pre.Bio.Data$Grade[i] <- "B"
    Pre.Bio.Data$GPA[i] <- 3.0
  }
  else if(G >= 2.6){
    Pre.Bio.Data$Grade[i] <- "B-"
    Pre.Bio.Data$GPA[i] <- 2.7
  }
  else if(G >= 2.2){
    Pre.Bio.Data$Grade[i] <- "C+"
    Pre.Bio.Data$GPA[i] <- 2.3
  }
  else if(G >= 1.9){
    Pre.Bio.Data$Grade[i] <- "C"
    Pre.Bio.Data$GPA[i] <- 2.0
  }
  else if(G >= 1.6){
    Pre.Bio.Data$Grade[i] <- "C-"
    Pre.Bio.Data$GPA[i] <- 1.7
  }
  else if(G >= 1.2){
    Pre.Bio.Data$Grade[i] <- "D+"
    Pre.Bio.Data$GPA[i] <- 1.3
  }
  else if(G >= 0.9){
    Pre.Bio.Data$Grade[i] <- "D"
    Pre.Bio.Data$GPA[i] <- 1.0
  }
  else if(G >= 0.6){
    Pre.Bio.Data$Grade[i] <- "D-"
    Pre.Bio.Data$GPA[i] <- 0.7
  }
  else{
    Pre.Bio.Data$Grade[i] <- "F"
    Pre.Bio.Data$GPA[i] <- 0.0
  }
}

```

```{r}
##Normalize Pre.Bio.Data Variables into useful object types
for(i in 1:length(Pre.Bio.Data$Unique_Student_ID)){
  Pre.Bio.Data$Unique_Student_ID[i] <- as.double(Pre.Bio.Data$Unique_Student_ID[i])
  Pre.Bio.Data$Unique_Instructor_ID[i] <- as.double(Pre.Bio.Data$Unique_Instructor_ID[i])
  Pre.Bio.Data$Year[i] <- as.double(Pre.Bio.Data$Year[i])
  Pre.Bio.Data$SUBJ_CODE[i] <- as.character(Pre.Bio.Data$SUBJ_CODE[i])
  Pre.Bio.Data$COURSE_NUMBER[i] <- as.double(Pre.Bio.Data$COURSE_NUMBER[i])
}

```
Issue with Data Set: Biology students stored seperately from math and chemistry students

```{r}

##Build Random ID assigner to normalize student and faculty naming with GenData

##Student Random ID Assigner

Pre.Bio.Data <- arrange(Pre.Bio.Data, `Unique_Student_ID`)

cntr <- 1
for(i in 1:length(Pre.Bio.Data$Unique_Student_ID)){
  
  if(i == 1){
    next.terms <- 0
    while(Pre.Bio.Data$Unique_Student_ID[i] == Pre.Bio.Data$Unique_Student_ID[i+next.terms]){
      next.terms <- next.terms + 1
    }
    next.terms <- next.terms - 1
    if((cntr %in% Gen_Data$`Student Random ID`) == FALSE){
      for(k in i: (i + next.terms)){
      Pre.Bio.Data$Unique_Student_ID[k] <- cntr
}
    }
    cntr <- cntr + 1
  }
  
    else if(Pre.Bio.Data$Unique_Student_ID[i] == Pre.Bio.Data$Unique_Student_ID[i-1]){
      ##Skips repeat student entry
      cntr <- cntr + 1
      
    }
    else{
     next.terms <- 0
      while(Pre.Bio.Data$Unique_Student_ID[i] == Pre.Bio.Data$Unique_Student_ID[i+next.terms]){
      next.terms <- next.terms + 1
    }
      next.terms <- next.terms - 1
    
       
      if((cntr %in% Gen_Data$`Student Random ID`) == FALSE){
        for(k in i: (i + next.terms)){
        Pre.Bio.Data$Unique_Student_ID[k] <- cntr
}
    }
      else{
        while(cntr %in% Gen_Data$`Student Random ID`){
        cntr <- cntr + 1
      }
        for(k in i: (i + next.terms)){
        Pre.Bio.Data$Unique_Student_ID[k] <- cntr
}
      
      }
      cntr <- cntr + 1
    }
    
  cntr < cntr + 1
}
```

```{r}

##Faculty Random ID Assigner
Pre.Bio.Data <- arrange(Pre.Bio.Data, `Unique_Instructor_ID`)

cntr <- as.double(1)
for(i in 1:length(Pre.Bio.Data$Unique_Instructor_ID)){
  
  if(i == 1){
    next.terms <- 0
    while(Pre.Bio.Data$Unique_Instructor_ID[i] == Pre.Bio.Data$Unique_Instructor_ID[i+next.terms]){
      next.terms <- next.terms + 1
    }
    next.terms <- next.terms - 1
    if((cntr %in% Gen_Data$`Faculty Random ID`) == FALSE){
      for(k in i: (i + next.terms)){
      Pre.Bio.Data$Unique_Instructor_ID[k] <- as.double(cntr)
}
    }
    cntr <- cntr + 1
  }
  
    else if(Pre.Bio.Data$Unique_Instructor_ID[i] == Pre.Bio.Data$Unique_Instructor_ID[i-1]){
      ##Skips repeat student entry
      cntr <- cntr + 1
      
    }
    else{
     next.terms <- 0
      while(Pre.Bio.Data$Unique_Instructor_ID[i] == Pre.Bio.Data$Unique_Instructor_ID[i+next.terms]){
      next.terms <- next.terms + 1
    }
      next.terms <- next.terms - 1
    
       
      if((cntr %in% Gen_Data$`Faculty Random ID`) == FALSE){
        for(k in i: (i + next.terms)){
        Pre.Bio.Data$Unique_Instructor_ID[k] <- as.double(cntr)
}
    }
      else{
        while(cntr %in% Gen_Data$`Faculty Random ID`){
        cntr <- cntr + 1
      }
        for(k in i: (i + next.terms)){
        Pre.Bio.Data$Unique_Instructor_ID[k] <- cntr
}
      
      }
      cntr <- cntr + 1
    }
    
  cntr < cntr + 1
}
```
Note: In Pre.Bio_Data, there were exactly 100 entries with term "Spring", out of 21538. There are no spring term entries for any of the chemistry nor math students in Gen_Data. In order to rectify this, I used a coin flip to split the spring term students between the winter and summer terms.
```{r}

##Term-inology converter; that's a pun :)
for(i in 1:length(Pre.Bio.Data$Term)){
  this.year <- Pre.Bio.Data$Year[i]
  if(Pre.Bio.Data$Term[i] == "Summer"){
    Pre.Bio.Data$Term[i] <- paste(this.year,"/","SU",sep = "")
  }
  if(Pre.Bio.Data$Term[i] == "Winter"){
    Pre.Bio.Data$Term[i] <- paste(this.year,"/","WI",sep = "")
  }
  if(Pre.Bio.Data$Term[i] == "Fall"){
    Pre.Bio.Data$Term[i] <- paste(this.year,"/","FA",sep = "")
  }
  if(Pre.Bio.Data$Term[i] == "Spring"){
    flip <- sample(0:1)
    if(flip == 0){
      Pre.Bio.Data$Term[i] <- paste(this.year,"/","WI",sep = "")
    }
    if(flip == 1){
    Pre.Bio.Data$Term[i] <- paste(this.year,"/","SU",sep = "")
    }
  }
}
rename(Pre.Bio.Data, Semester = Term, `Student Random ID` = Unique_Student_ID, `Faculty Random ID` = Unique_Instructor_ID, )


```

```{r}
Bio.Data <- data.frame(`Student Random ID` = Pre.Bio.Data$Unique_Student_ID, `Course Code` = Pre.Bio.Data$COURSE_NUMBER, `Grade` = Pre.Bio.Data$Grade, `Faculty Random ID` = Pre.Bio.Data$Unique_Instructor_ID,  `Semester` = paste(Pre.Bio.Data$Year, "/", Pre.Bio.Data$Term, sep = ""), `GPA` = Pre.Bio.Data$GPA)


```